Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-19 mpaper/paper.tex
Restructuring + importing the related work section

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage{thmtools, thm-restate}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{layouts}
\usepackage{xcolor}

@@

\section{Introduction}
% TODO: rewrite and make it good
% 1. Problem (with examples, intuitive)
% 2. Its importance/relevance/motivation
% 3. My idea
% 4. List of contributions
% Forward-reference which section deals with what

@@

\begin{definition}[Norms]
  For any finite-dimensional vector $\mathbf{x} = (x_1, \dots, x_n)^\intercal$,
  its \emph{maximum norm} is
  \[
    \lVert \mathbf{x} \rVert_\infty = \max_i |x_i|
  \]
  whereas its \emph{taxicab} (or \emph{Manhattan}) \emph{norm} is
  \[
    \lVert \mathbf{x} \rVert_1 = \sum_{i = 1}^n |x_i|.
  \]
  Let $\mathbf{A}$ be a matrix. For any vector norm $\lVert
  \cdot \rVert_p$, we can also define its \emph{induced norm} for matrices as
  \[
    \lVert \mathbf{A} \rVert_p = \sup_{\mathbf{x} \ne \mathbf{0}} \frac{\lVert
      \mathbf{Ax} \rVert_p}{\lVert \mathbf{x} \rVert_p}.
  \]
  In particular, for $p = \infty$, we have
  \[
    \lVert \mathbf{A} \rVert_\infty = \max_i \sum_{j} |A_{i,j}|.
  \]
\end{definition}

@@

\section{Solution}
% TODO: give an overview of this section. Also, consider renaming it or
% splitting it into multiple sections

For any matrix $\mathbf{A}$, we will use either $A_{i,j}$ or
$[\mathbf{A}]_{i,j}$ to denote the element of $\mathbf{A}$ in row $i$ and column
$j$. Moreover, we use $\tr(\mathbf{A})$ to denote its \emph{trace} and
$\adj(\mathbf{A})$ for its \emph{adjugate} (or \emph{classical adjoint}). For
any vector $\mathbf{x}$, we write $\mathbb{R}_d[\mathbf{x}]$ to denote a vector
space of polynomials with degree at most $d$, where variables are elements of
$\mathbf{x}$, and coefficients are in $\mathbb{R}$.

@@

\subsection{Details of the Model}

@@

\subsection{Theoretical Justification}

@@

In order to fully understand the model's behaviour, we focus on a three-state
MDP where the agent can deterministically move from any state to any other
state. More formally, we set $\mathcal{S} = \{ s_1, s_2, s_3 \}$, $\mathcal{A} =
\{ a_1, a_2 \}$,
\begin{align*}
  &\mathcal{T}(s_1, a_1, s_2) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_2, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_3, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_2) = 1,
\end{align*}
all other values of $\mathcal{T}$ to zero, and $\gamma = 0.9$. We
also set the inducing points to be equal to the three states in $\mathcal{S}$,
add a single feature $f : \mathcal{S} \to \mathbb{R}$ such that
\[
  f(s_1) = 1, \quad f(s_2) = 2, \quad f(s_3) = 3,
\]
and create two demonstrations: $\zeta_1 = \{ (s_1, a_1) \}$ and $\zeta_2 = \{
(s_3, a_2) \}$. In other words, the demonstrations move from $s_1$ and $s_3$ to
$s_2$. Therefore, we would expect the reward (and value) of $s_2$ to be the
highest.

\begin{figure*}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{sage1}
    \includegraphics[width=\columnwidth]{sage2}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{elbo_over_B_short}
  \end{minipage}
  \caption{How a well-defined problem with a correct solutions becomes a
    well-defined problem with an incorrect solution. In each plot, the function
    we are trying to optimise is in blue, and its derivative is in red. The
    plots on the left are for $q(\mathbf{u})$ and its derivatives with respect
    to $B_{1,1}$ (at the top) and $B_{2,1}$ (at the bottom), whereas the plots
    on the right are for $\mathbb{E}[v]$ and its corresponding derivatives.}
\end{figure*}

TODO: describe the problem with B.

\begin{figure*}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{convergence1}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=\columnwidth]{convergence2}
  \end{minipage}
  \caption{The convergence of $\mathcal{L}$ (at the top) and several example
    policies (at the bottom) over a number of iterations for Scenario~1 on the
    left and Scenario~2 on the right.}
  \label{fig:convergence1}
\end{figure*}

@@

Because of this issue, we look at the convergence of our model using a constant
step size in two different scenarios:
\begin{itemize}
\item In \textbf{Scenario 1}, we remove $\frac{\partial \mathbb{E}[v]}{\partial
    \mathbf{B}}$ from $\frac{\partial \mathcal{L}}{\partial \mathbf{B}}$. This
  essentially optimises $\mathbf{B}$ to match $\Kuu$, because then
  $\frac{\partial \mathcal{L}}{\partial \mathbf{B}}$ becomes
  $-\frac{\partial}{\partial \mathbf{B}} \DKL{q(\mathbf{u})}{p(\mathbf{u})}$,
  i.e., we are optimising $\mathbf{B}$ to minimise the difference between the
  prior and the posterior of $\mathbf{u}$.
\item In \textbf{Scenario 2}, we set $\mathbf{B} = \mathbf{I}_m$, and do not
  optimise it at all.
\end{itemize}
% TODO: use \ell when introducing the norms

@@

\section{Related Work}
% TODO: comprehensive literature survey

@@

>>>>

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage{thmtools, thm-restate}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{layouts}
\usepackage{xcolor}
\usepackage{booktabs}

@@

\section{Introduction} % first page
% TODO: rewrite and make it good
% 1. Problem (with examples, intuitive)
% 2. Its importance/relevance/motivation
% 3. My idea
% 4. List of contributions
% Forward-reference which section deals with what

@@

\begin{definition}[Norms]
  For any finite-dimensional vector $\mathbf{x} = (x_1, \dots, x_n)^\intercal$,
  its \emph{maximum norm} (\emph{$\ell_\infty$-norm}) is
  \[
    \lVert \mathbf{x} \rVert_\infty = \max_i |x_i|
  \]
  whereas its \emph{$\ell_1$-norm} is
  \[
    \lVert \mathbf{x} \rVert_1 = \sum_{i = 1}^n |x_i|.
  \]
  Let $\mathbf{A}$ be a matrix. For any vector norm $\lVert
  \cdot \rVert_p$, we can also define its \emph{induced norm} for matrices as
  \[
    \lVert \mathbf{A} \rVert_p = \sup_{\mathbf{x} \ne \mathbf{0}} \frac{\lVert
      \mathbf{Ax} \rVert_p}{\lVert \mathbf{x} \rVert_p}.
  \]
  In particular, for $p = \infty$, we have
  \[
    \lVert \mathbf{A} \rVert_\infty = \max_i \sum_{j} |A_{i,j}|.
  \]
\end{definition}

@@

\section{The Model}

% TODO: make this fit the narrative
For any matrix $\mathbf{A}$, we will use either $A_{i,j}$ or
$[\mathbf{A}]_{i,j}$ to denote the element of $\mathbf{A}$ in row $i$ and column
$j$. Moreover, we use $\tr(\mathbf{A})$ to denote its \emph{trace} and
$\adj(\mathbf{A})$ for its \emph{adjugate} (or \emph{classical adjoint}). For
any vector $\mathbf{x}$, we write $\mathbb{R}_d[\mathbf{x}]$ to denote a vector
space of polynomials with degree at most $d$, where variables are elements of
$\mathbf{x}$, and coefficients are in $\mathbb{R}$.

@@

@@

\section{Theoretical Justification}

@@

In order to fully understand the model's behaviour, we focus on a three-state
MDP where the agent can deterministically move from any state to any other
state. More formally, we set $\mathcal{S} = \{ s_1, s_2, s_3 \}$, $\mathcal{A} =
\{ a_1, a_2 \}$,
\begin{align*}
  &\mathcal{T}(s_1, a_1, s_2) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_2, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_3, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_2) = 1,
\end{align*}
all other values of $\mathcal{T}$ to zero, and $\gamma = 0.9$. We
also set the inducing points to be equal to the three states in $\mathcal{S}$,
add a single feature $f : \mathcal{S} \to \mathbb{R}$ such that
\[
  f(s_1) = 1, \quad f(s_2) = 2, \quad f(s_3) = 3,
\]
and create two demonstrations $\zeta_1 = \{ (s_1, a_1) \}$ and $\zeta_2 = \{
(s_3, a_2) \}$ that correspond to moving from $s_1$ and $s_3$ to $s_2$.
Therefore, we would expect the reward of $s_2$ to be higher than the other two
rewards in order to reflect this.

\begin{figure*}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.8\columnwidth]{sage1}
    \includegraphics[width=0.8\columnwidth]{sage2}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.8\columnwidth]{elbo_over_B_short}
  \end{minipage}
  \caption{How a well-defined problem with a correct solutions becomes a
    well-defined problem with an incorrect solution. In each plot, the function
    we are trying to optimise is in blue, and its derivative is in red. The
    plots on the left are for $q(\mathbf{u})$ and its derivatives with respect
    to $B_{1,1}$ (at the top) and $B_{2,1}$ (at the bottom), whereas the plots
    on the right are for $\mathbb{E}[v]$ and its corresponding derivatives.}
  \label{fig:B_gradient}
\end{figure*}

\begin{figure*}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.8\columnwidth]{convergence1}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.8\columnwidth]{convergence2}
  \end{minipage}
  \caption{The convergence of $\mathcal{L}$ (at the top) and several example
    policies (at the bottom) over a number of iterations for Scenario~1 on the
    left and Scenario~2 on the right.}
  \label{fig:convergence1}
\end{figure*}

@@

Unfortunately, we are not able to use $\frac{\partial \mathcal{L}}{\partial
  \mathbf{B}}$ in order to optimise $\mathbf{B}$. We illustrate the problem in
Figure \ref{fig:B_gradient}. On the left side of the figure, we plot how
$q(\mathbf{u})$ behaves as a function of a diagonal and a non-diagonal element
of $\mathbf{B}$. Both functions have maximum values that can be attained by
following the corresponding derivatives. However, when these derivatives are
used to estimate $\frac{\partial \mathbb{E}[v]}{\partial \mathbf{B}}$, the
resulting derivatives no longer match their corresponding functions, although
the functions themselves still have optimal values: at or below $1$ for
$B_{1,1}$ and at $0$ for $B_{2,1}$. This leads us to consider two restrictions
of the model in order to investigate its convergence behaviour:
\begin{itemize}
\item In \textbf{Scenario 1}, we remove $\frac{\partial \mathbb{E}[v]}{\partial
    \mathbf{B}}$ from $\frac{\partial \mathcal{L}}{\partial \mathbf{B}}$. This
  essentially optimises $\mathbf{B}$ to match $\Kuu$, because then
  $\frac{\partial \mathcal{L}}{\partial \mathbf{B}}$ becomes
  $-\frac{\partial}{\partial \mathbf{B}} \DKL{q(\mathbf{u})}{p(\mathbf{u})}$,
  i.e., we are optimising $\mathbf{B}$ to minimise the difference between the
  prior and the posterior of $\mathbf{u}$.
\item In \textbf{Scenario 2}, we set $\mathbf{B} = \mathbf{I}_m$, and do not
  optimise it at all.
\end{itemize}

@@

\section{Related Work} % 0.5-1 page
% TODO: rewrite and make this section smaller

As mentioned in the introduction, most IRL algorithms assume that the reward
function can be represented as a linear combination of features. This assumption
originated in one of the earliest papers on the topic by Ng and Russell
\cite{DBLP:conf/icml/NgR00}, which introduced several linear programming
approaches to the problem. The authors also noticed that often
multiple reward functions can explain the same behaviour, and suggested
heuristics for reward functions that are `far away' from reward functions that
do not fit the data.

A few years later, Abbeel and Ng \cite{DBLP:conf/icml/PieterN04} developed an
algorithm for the same formulation of the problem with a guarantee to converge
quickly. Neu and Szepesv\'ari \cite{DBLP:conf/uai/NeuS07} identified a weakness
in Abbeel and Ng's approach: the algorithm requires features to be
`appropriately' scaled, and optimal scaling may not be known. Instead, they
suggest a way to combine IRL with \emph{apprenticeship learning}, i.e., a
supervised learning task for optimal \emph{policy} recovery (whereas IRL focuses
on recovering the reward function).

Ramachandran and Amir \cite{DBLP:conf/ijcai/RamachandranA07} were the first to
formulate IRL in terms of Bayesian learning. While the model is easily
interpretable and able to handle experts that make mistakes, the algorithm can
only handle small state spaces and requires Monte Carlo Markov Chain (MCMC)
sampling for inference.

Ziebart et al. \cite{ziebart2008maximum} keep the linearity assumption, but
introduce an influential idea: resolving the ambiguity when multiple reward
functions explain the data by appealing to the maximum entropy principle.

Choi and Kim \cite{DBLP:conf/ijcai/ChoiK13} extend the Bayesian model to learn
good features as well as the reward function, trying to overcome the limitation
of linearity. However, the approach is quite limiting: all features are assumed
to have Boolean values, and the algorithm simply learns their conjunctions.

Levine et al. \cite{DBLP:conf/nips/LevinePK11} are the first to suggest a way to
learn nonlinear reward functions without harsh restrictions on the problem
domain by using GPs. We base our work primarily on their paper, and the
weaknesses we hope to address have already been covered in the previous section.
A recent extension to their work by Jin et al. \cite{DBLP:conf/uai/JinDAS17}
aims to harness the power of deep learning by using several layers of GPs,
making the model less dependent on being provided good features. They also use
VI, but with a few simplifying assumptions: deterministic training conditional
for the reward vector, and fully independent training conditional for the latent
state.

Finally, instead of using GPs to model nonlinear reward functions, one can use a
neural network (NN), as demonstrated by Wulfmeier et al.
\cite{wulfmeier2015maximum}. Their approach benefits from constant time
inference and the ability to learn complex features either from already-given
features or even from raw data. The only disadvantage (as demonstrated in the
paper) is that NNs converge slower than GPs (as measured by the number of
demonstrations).

\subsection{Variational Inference}

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as NNs)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, with
many methods providing a time complexity of $\mathcal{O}(m^3)$. Below we provide
a short overview of various assumptions that have been used in approximating
\emph{sparse} GPs (i.e., GPs that use inducing points), following on a paper by
Qui\~{n}onero-Candela and Rasmussen \cite{DBLP:journals/jmlr/CandelaR05}.

Table ?? further summarises some of the recent and/or
influential GP approximation approaches that might be relevant to our situation.
For many of the variables in our model (including the inducing points and the
hyperparameters of the covariance function), we have three possible ways of
handling them:
\begin{itemize}
\item Provide a full Bayesian treatment by defining a prior probability
  distribution. While this option can easily prevent overfitting, the choice of
  prior can be difficult to justify.
\item Fix the value of the variable. We can then optimise the value outside of
  the main algorithm, e.g., by running the algorithm with a set of possible
  values and keeping the value that produces the best performance. This option
  can be slow if we are interested in finding the optimal value, but could be
  appropriate if the variable has little impact on the algorithm's performance.
\item Treat the variable as a variational parameter. This method will
  efficiently optimise the value of the variable, but is vulnerable to
  overfitting. Furthermore, this requires us to find the derivative of the ELBO
  w.r.t. the variable, which could be difficult.
\end{itemize}

Titsias \cite{DBLP:journals/jmlr/Titsias09} is the first to suggest a
variational approximation where inducing points are treated as variational
parameters. The paper has influenced many of the later works covered in this
section, and has been reformulated by Gal et al. \cite{DBLP:conf/nips/GalWR14}
into a distributed algorithm. We will fix the inducing points for the baseline
version of our algorithm, but keep the idea in mind for possible future work.

Hensman et al. \cite{DBLP:conf/uai/HensmanFL13} reduce the complexity from
$\mathcal{O}(nm^2)$ to $\mathcal{O}(m^3)$ using a stochastic variational
inference (SVI) approach. While results such as efficiently computable natural
gradients and analytically tractable optimal solutions for variational
parameters are unlikely to transfer to a different problem domain, the overall
SVI framework and an approximating Gaussian distribution for $\mathbf{u}$ will
play an important role in our proposal.

Cheng and Boots \cite{DBLP:conf/nips/ChengB17} suggest using different bases in
the reproducing kernel Hilbert space for the mean and covariance functions.
The paper makes few assumptions and mostly relies on the work of
Titsias \cite{DBLP:journals/jmlr/Titsias09}, making it seem transferable to a
new domain. However, applying the idea to IRL is likely to be beyond the scope
of the project.

The variational Fourier features (VFF) algorithm by Hensman et al.
\cite{DBLP:journals/jmlr/HensmanDS17} is only defined for Mat\'ern kernels,
which is likely to be too restrictive for our situation (the original paper on
using GPs for IRL \cite{DBLP:conf/nips/LevinePK11} used the automatic relevance
detection kernel that has weights controlling how important each feature is).
While extending VFF to support a flexible class of kernels defined by Wilson and
Adams \cite{pmlr-v28-wilson13} is an interesting and promising avenue of work,
it is likely to be beyond the scope of this project as well.

While Peng et al. \cite{DBLP:conf/icml/PengZZQ17} provide a highly efficient
distributed implementation, the derivation of the ELBO relies primarily on the
fact that the evidence for a GP is a Gaussian, while in our case the evidence is
anything but. Hence, it is unlikely that the ideas from this paper could be
applicable to our IRL model.

Finally, since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing a wide range of possible probability
distributions. The primary way of representing complex posteriors in VI is by
using \emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Normalising flows, along
with some of the ideas mentioned previously, could provide great benefits to the
GP IRL VI model. However, we are forced to keep any non-essential features as
possible future work in order to make the project feasible in a given time
frame. Furthermore, note that due to various differences between regression and
IRL, we cannot simply apply a GP VI model as a whole. Instead, smaller tricks
and ideas from various papers can (and will) be used throughout the project.

@@

%Why doesn't it work? Perhaps a stupid bug.
